Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-05-01 sources/bmc_template_submission/bmc_article.tex
last edits

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides
the opportunity to simultaneously study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across biological databases.
Physcraper's main goal is to use its interoperability framework to build upon published alignments and phylogenies
to extend existing phylogenetic inferences with more data and taxa, which improve our
phylogenetic reconstructions \cite{hillis1996inferring, natsidis2019phylogenomics},
time of divergence estimates \cite{schulte2013undersampling, soares2015influence},
biogeographic analyses \cite{kayaalp2017back},
and help in resolving phylogenetic conflict \cite{hedtke2006resolution, townsend2010optimal, natsidis2019phylogenomics}.
Physcraper updates a starting single locus alignment and corresponding phylogeny
with public DNA data from GenBank \cite{benson2000genbank}, and links tip taxon labels in the updated trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}, streamlining connections across biological databases.

Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public biological databases constitute an amazing
resource for scientific discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of data across biological databases are varying taxonomic idiosyncrasies.
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for automatic taxonomic name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020},
and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019}, among many others.
All of OpenTree tools and methods are implemented via OpenTree's Application Programming Interfaces (APIs),
as open access services for use by the general public \cite{opentreeAPIs}.
Physcraper leverages on existing OpenTree's taxonomic APIs to
standardize taxon names in any phylogeny, providing
a framework for automatic connection of updated phylogenies with data from any biological database.

Decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that have the potential to be used for phylogenetic
reconstruction at many scales \cite{sanderson2008phylota}.
More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
However, even thirteen years later, phylogenetic estimates for most of these taxa
are still not available \cite{mctavish_how_2017}.
Using its unified taxonomy, OpenTree assembles a comprehensive synthetic tree of life comprising
2.3 million tips, of which around only
90,000 are supported by publicly available expert phylogenetic data, uploaded to OpenTrees' database (the Phylesystem
\cite{mctavish2015phylesystem}) by human curators - the remaining 1.4 million taxa are placed
in the synthetic tree based on taxonomy.
There is a considerable amount of phylogenetically
informative data in GenBank with the potential to fill these phylogenetic gaps in
the tree of life, but this data either has not been analysed or the analyses
have not been made publicly available \cite{mctavish_how_2017}.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides
the opportunity to simultaneously study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across biological databases.
Physcraper's main goal is to use its interoperability framework to build upon published alignments and phylogenies
to extend existing phylogenetic inferences with more data and taxa, which improve our
phylogenetic reconstructions \cite{hillis1996inferring, natsidis2019phylogenomics},
time of divergence estimates \cite{schulte2013undersampling, soares2015influence},
and biogeographic analyses \cite{kayaalp2017back},
and help in resolving phylogenetic conflict \cite{hedtke2006resolution, townsend2010optimal, natsidis2019phylogenomics}.
Physcraper updates a starting single locus alignment and corresponding phylogeny
with public DNA data from GenBank \cite{benson2000genbank}, and links tip taxon labels in the updated trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}, streamlining connections between phylogenies and biological data.
Physcraper links tip taxon labels in the updated trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}, eefectively streamlining connections between phylogenies and biological data.

Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public biological databases constitute an amazing
resource for scientific discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of data across biological databases are varying taxonomic idiosyncrasies.
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for automatic taxonomic name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020},
and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019}, among many others.
All of OpenTree tools and methods are implemented via OpenTree's Application Programming Interfaces (APIs),
as open access services for the general public \cite{opentreeAPIs}.
Physcraper leverages on existing OpenTree's taxonomic APIs to
standardize taxon names in any phylogeny, providing
a framework for automatic connection of updated phylogenies with data from any biological database.

Decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that have the potential to be used for phylogenetic
reconstruction at many scales \cite{sanderson2008phylota}.
More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
However, even thirteen years later, phylogenetic estimates for most of these taxa
are still not available \cite{mctavish_how_2017}.
OpenTree assembles a comprehensive synthetic tree of life comprising
2.3 million tips, of which around
90,000 are supported by publicly available expert phylogenetic data, uploaded to OpenTrees' database (the Phylesystem
\cite{mctavish2015phylesystem}) by volunteer curators - the remaining 1.4 million taxa are placed
in the synthetic tree based on OpenTree's unified taxonomy.
There is a considerable amount of phylogenetically
informative data in GenBank with the potential to fill these phylogenetic gaps in
the tree of life, but this data either has not been analysed or the analyses
have not been made publicly available and accesible \cite{mctavish_how_2017}.
